Tidy Data

### 2014-01-14 tidy-data.tex
Tidy Data

\begin{enumerate}
  \item Each variable forms a column,
  \item Each observation forms a row,
  \item Each type of observational unit forms a table.
\end{enumerate}

\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus on a single dataset, not the many connected datasets common in relational databases. \textbf{Messy data} is any other other arrangement of the data.

Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1} because each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable. We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.

Tidy data makes it easy for an analyst or a computer to extract needed variables because it provides a standard way of structuring a dataset. Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables. This slows analysis and invites errors. If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a standard, simple way. Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.

>>>>

\begin{enumerate}
  \item Each variable forms a column.
  \item Each observation forms a row.
  \item Each type of observational unit forms a table.
\end{enumerate}

\noindent This is Codd's 3rd normal form \citep{codd:1990}, but with the constraints framed in statistical language, and the focus put on a single dataset rather than the many connected datasets common in relational databases. \textbf{Messy data} is any other other arrangement of the data.

Table~\ref{tbl:preg-tidy} is the tidy version of Table~\ref{tbl:preg-raw-1}. Each row represents an observation, the \code{result} of one \code{treatment} on one \code{person}, and each column is a variable. We could also choose to drop the row containing the missing value, because missing values should only be omitted if they are structural, impossible because of the design of the experiment.

Because it provides a standard way of structuring a dataset, tidy data makes it easy for an analyst or a computer to extract needed variables. Contrast Table~\ref{tbl:preg-tidy} to Table~\ref{tbl:preg-raw-1}: in Table~\ref{tbl:preg-raw-1} you need to use different strategies to extract different variables. This slows analysis and invites errors. If you consider how many data analysis operations involve all of the values in a variable (every aggregating function), you can see how important it is to extract these values in a standard, simple way. Tidy data is particularly well suited for vectorised programming languages like \proglang{R}, because the layout ensures that values of different variables from the same observation are always paired.
